Static Race Detection in OS Kernels by Mining Locking Rules

### 2023-08-02 1.introduction.tex
Finish a temporary version (without comparing experiment and some evaluation results)

To take advantage of multiple-core architecture of modern CPUs, an OS kernel is 
expected to handle concurrency. However, concurrent execution can inevitably 
introduce concurrency issues. Data race is one of the most harmful concurrency 
issues that occurs when multiple threads access the same variable concurrently 
and at least one of the accesses is a write. Some data races are quite 
dangerous and can cause serious bugs such as null-pointer dereferences, data 
inconsistency and infinite loops.

To detect data races effectively, some approaches~\cite{Boyapati:OOPSLA02, 
Anderson:PLDI08, Anderson:PLDI09, Zhou:MICRO19, Flanagan:PASTE01, 
Flanagan:PLDI00, Sadowski:PLATEAU14, ClangThreadSafety, Blackshear:OOPSLA18} 
rely on developers to supply annotations that describe the locking discipline 
such as which lock protects a specific variable, and detect data races by 
exploiting variable accesses that violate these disciplines. For example, Clang 
thread safety analysis~\cite{ClangThreadSafety} requires developers to label a 
variable and the lock to protect it with a {\tt GUARDED\_BY} attribute. And 
then detects data races by finding variable accesses that violate the given 
attribute. However, such annotation-based approaches are not suitable for race 
detection in OS kernels, because even an expert developer can not provide 
accurate annotations, due to poor documentation and complicated logic of OS 
codes.

Some static approaches~\cite{Choi:PLDI02, Engler:SOSP03, Voung:FSE07, 
Pratikakis:PLDI06, Naik:PLDI06} employ lockset-based analysis to detect data 
races automatically. They first collect locks that are acquired when a given 
variable is accessed, and then mine locking rules about whether a given 
variable should be protected by a specific lock. At last, they detect data 
races by checking whether variable accesses violate the mined locking rules. 
However, they do not consider alias relations~\cite{Voung:FSE07, Engler:SOSP03} 
or just use imprecise flow-insensitive alias analysis~\cite{Choi:PLDI02, 
Pratikakis:PLDI06, Naik:PLDI06}, and thus can introduce both false positives 
and false negatives.

Dynamic analysis can acquire run-time information such as memory address, and 
thus complex alias analysis can be avoided. Therefore, some 
approaches~\cite{Lochmann:EuroSys19, Lu:SOSP07, Lu:FSE18, Joshi:ASE08, 
Liu:NSDI07} detect data races dynamically to improve precision. They mine 
implicit code rules in softwares by statistically analyzing execution traces, 
and then use the mined rules to detect related bugs. For example, 
LockDoc~\cite{Lochmann:EuroSys19} records accesses to variables and lock 
acquisitions of an instrumented Linux kernel, and then infers locking rules 
from the recorded accesses. After that, LockDoc automatically locates variables 
access that violate the inferred locking rules to detect concurrency bugs such 
as data races. However, dynamic analysis suffers from low code coverage, and 
thus the locking rules inferred through execution traces can be imprecise. 
Besides, dynamic analyses are hard to deploy, because they need to run OS 
kernels with instruments. At last, it is difficult for dynamic tools to 
estimate the harmfulness of a data race, because race conditions are hard to 
trigger by running the checked software~\cite{Fonseca:DSN10, 
Burckhardt:ASPLOS10, Liu:FSE14, Zhou:EASE15}.

In this paper, we design a practical static analysis approach named \sys, 
to automatically detect data races and estimate the harmfulness of these data 
races in OS kernels effectively. \sys consists of three key techniques:

(T1) \sys uses an {\em alias-aware rule mining method} to automatically 
deduce locking rules. We observe that an access to a variable is often 
protected by the lock in the same data structure as the accessed field. And 
this relationship between accessed field and the protecting lock can be 
effectively inferred from a special data structure named alias 
graph~\cite{Li:ASPLOS22, Kastrinis:CC18}. Specifically, fields in the same data 
structure have the same ancestor in the alias graph, and thus whether an 
accessed field and a specific lock are in the same data structure can be 
inferred by finding a common ancestor in the alias graph. Moreover, with 
benefits from precise field-sensitive alias relationships of alias graph, 
\sys can effectively extract the accessed field and its corresponding 
protecting lock. After obtaining the accessed field and protecting lock pairs, 
\sys calculates the proportion of field accesses protected by a specific 
lock in all field accesses. If the proportion is larger than a given threshold, 
the accessed field is determined to be protected by the protecting lock.

(T2) \sys uses a {\em lock-usage analysis} to filter out false positives 
validating concurrency of code paths. After mining locking rules, \sys detects 
data races by checking whether a given variable access violates the 
locking rules. To reduce false negatives, \sys conservatively assumes 
every two functions can be executed concurrently, and thus can introduce false 
positives. However, we observe that each kernel module has an initialization 
phase, after which many functions can be called concurrently by other modules. 
When performing initialization, the kernel module serially initialized the lock 
and prepare other data for subsequent operations. And thus functions with the 
lock initialization and functions called by them tend not to be executed 
concurrently. Based on this observation, \sys extracts all functions 
reachable from the function with a lock initialization operation such as {\tt 
spin\_lock\_init()}, and suppose that this function can not be executed in 
parallel.

(T3) \sys uses a {\em pattern-based estimation} to extract harmful races 
that can trigger memory or logical bugs such as null-pointer dereferences and 
data inconsistency. Some data races are benign, and developers do not put 
effort into repairing them because they can not cause serious memory or logical 
bugs. Therefore, it is important to estimate the harm a data race can bring. We 
observe that harmful data races can be estimated by given patterns. For 
example, if a raced data performs as an operand of a dereference instruction, a 
null-pointer dereference can occur; if a raced data is accessed for more than 
once and data inconsistency can occur. Based on this observation, \sys 
uses some patterns to extract harmful races that can cause memory or logical 
bugs automatically.

@@

\begin{itemize}
	\item We analyze the challenges of data race detection in OS kernels, and 
	propose three key techniques to address	these challenges: (T1) an {\em 
	alias-aware rule mining method} to automatically deduce locking rules; (T2) 
	a {\em lock-usage analysis} to filter out false positives caused by 
	concurrency; (T3) a {\em pattern-based estimation} to extract harmful data 
	races that can trigger memory or logical bugs such as null-pointer 
	dereferences and data inconsistency.
	\item Based on these three key techniques, we design a practical static 
	analysis approach named \sys, to effectively detect data races and 
	estimate the harmfulness of these data races in OS kernels.
	\item We have evaluated \sys on Linux 6.2, and find xxx data races, 	
	with a false positive rate of xxx. Among these data races, xxx are 
	estimated to be harmful. We have reported these harmful bugs to Linux 
	kernel developers, and xxx of them have been confirmed.
\end{itemize}

>>>>

To take advantage of multiple-core architecture of modern CPUs, an OS kernel is 
expected to handle concurrency. However, concurrent execution can inevitably 
introduce concurrency issues. Data race is one of the most harmful concurrency 
issues that occurs when multiple threads access a shared variable concurrently 
and at least one of the accesses is a write. Some races are harmful and can 
cause serious bugs like null-pointer dereferences, data inconsistencies and 
double fetches.

To detect data races effectively, some approaches~\cite{Boyapati:OOPSLA02, 
Anderson:PLDI08, Anderson:PLDI09, Zhou:MICRO19, Flanagan:PASTE01, 
Flanagan:PLDI00, Sadowski:PLATEAU14, ClangThreadSafety, Blackshear:OOPSLA18} 
rely on developers to supply annotations that describe the locking discipline 
such as which lock is required for accesses to a specific variable, and detect 
data races by searching variable accesses that violate these disciplines. For 
example, Clang thread safety analysis~\cite{ClangThreadSafety} requires 
developers to label a variable and the lock to protect it with a {\tt 
GUARDED\_BY} attribute. And then detects data races by finding variable 
accesses that violate the given attribute. However, such annotation-based 
approaches are not suitable for race detection in OS kernels, because even an 
expert developer can not provide accurate annotations, due to poor document and 
complicated logic of OS codes.

Other static approaches~\cite{Choi:PLDI02, Engler:SOSP03, Voung:FSE07, 
Pratikakis:PLDI06, Naik:PLDI06} employ lockset-based analysis to detect data 
races automatically. For a given variable, they first collect locks that are 
acquired for each access to it, and then detect data races by checking whether 
the intersection of locks for different accesses is empty. However, they do not 
consider alias relationships~\cite{Voung:FSE07, Engler:SOSP03} or just use 
imprecise flow-insensitive alias analysis~\cite{Choi:PLDI02, Pratikakis:PLDI06, 
Naik:PLDI06}, and thus can introduce both false positives and negatives.

Dynamic analysis can acquire run-time information such as memory address, and 
thus complex alias analysis can be avoided. Therefore, some 
approaches~\cite{Lochmann:EuroSys19, Lu:SOSP07, Lu:FSE18, Joshi:ASE08, 
Liu:NSDI07} detect data races dynamically to improve precision. They mine 
implicit code rules in softwares by statistically analyzing execution traces, 
and then use the mined rules to detect related bugs. For example, 
LockDoc~\cite{Lochmann:EuroSys19} records accesses to variables and lock 
acquisitions of an instrumented Linux kernel, and then infers locking rules 
from the recorded accesses. After that, LockDoc automatically locates variable 
accesses that violate the inferred locking rules to detect concurrency issues 
such as data races. However, dynamic analysis suffers from low code coverage, 
and thus the locking rules inferred through execution traces can be imprecise. 
Besides, dynamic analyses are hard to deploy, because they need to run OS 
kernels with instruments. At last, it is difficult for dynamic tools to 
estimate the harmfulness of a data race, because race conditions are hard to 
trigger by running the checked software~\cite{Fonseca:DSN10, 
Burckhardt:ASPLOS10, Liu:FSE14, Zhou:EASE15}.

In this paper, we design a practical static analysis approach named \sys, 
to automatically detect data races and estimate the harmfulness of these data 
races in OS kernels. \sys consists of three key techniques:

(T1) \sys uses an {\em alias-aware rule mining method} to automatically deduce 
locking rules about whether accesses to a specific data structure field should 
be protected and which data structure field the protecting lock exist in. We 
observe that the variable that is accessed and the lock to protect the access 
often exist in the same data structure. And this relationship between the 
accessed variable and the protecting lock can be effectively inferred from a 
special data structure named alias graph~\cite{Li:ASPLOS22, Kastrinis:CC18}. 
Specifically, variables in the same data structure have a common ancestor in 
the alias graph, and thus whether an accessed variable and a specific lock are 
in the same data structure can be inferred through finding a common ancestor in 
the alias graph. Moreover, with benefits from precise field-sensitive alias 
relationships of alias graph, \sys can effectively extract the accessed field 
and its corresponding lock field. After obtaining the accessed field and lock 
field pairs, \sys calculates the proportion of field accesses protected by a 
specific lock field in all field accesses. If the proportion is larger than a 
given threshold, the accessed field is determined to be protected by the lock 
field for a specific data structure.

(T2) \sys uses a {\em lock-usage analysis} to filter out false positives by 
validating concurrency of the kernel code. After mining locking rules, \sys 
detects data races by checking whether a given variable access violates the 
locking rules. To reduce false negatives, \sys conservatively assumes every two 
functions can execute concurrently, and thus can introduce false positives. 
However, we observe that each kernel module has an initialization phase, after 
which many functions can be called concurrently by other modules. When 
performing initialization, the kernel module serially initializes the locks and 
prepares other data for subsequent operations. And thus functions with the lock 
initialization and functions called by them tend not to execute concurrently. 
Based on this observation, \sys extracts all functions that are reachable from 
lock initialization functions such as {\tt spin\_lock\_init()} in the function 
call graph, and suppose that these functions can not execute concurrently.

(T3) \sys uses a {\em pattern-based estimation} to extract harmful races that 
can trigger memory or logic bugs such as null-pointer dereferences, data 
inconsistencies and double fetches. Some data races are benign, and developers 
do not put effort into repairing them because they can not cause serious memory 
or logic bugs. Therefore, it is important to estimate the harmfulness a data 
race can bring. We observe that harmful data races can be estimated by specific 
patterns. For example, if a raced data performs as an operand of a dereference 
instruction, a null-pointer dereference can occur; if a raced data structure is 
accessed for more than once and each access gets its different fields, a data 
inconsistency can occur; if a data race is caused by an unprotected write, and 
it is likely to introduce double fetches.Based on this observation, \sys 
exploits some patterns to extract harmful data races that can cause memory or 
logic bugs automatically.

@@

\begin{itemize}
	\item We analyze the challenges to detect races in kernels, and propose 
	three key techniques to address	these challenges: (T1) an {\em alias-aware 
	rule mining method} to automatically deduce locking rules; (T2) a {\em 
	lock-usage analysis} to filter out false positives caused by code that can 
	not execute concurrently; (T3) a {\em pattern-based estimation} to extract 
	harmful data races that can cause memory or logic bugs such as null-pointer 
	dereferences, data inconsistencies and double fetches.		
	\item Based on these three key techniques, we design a practical static 	
	analysis approach named \sys, to effectively detect data races and 	
	estimate the harmfulness of these data races in OS kernels.
	\item We have evaluated \sys on Linux 6.2, and find 273 real data races, 	
	with a false positive rate of 19.9\%. Among these data races, 88 are 
	estimated to be harmful. We have reported these harmful bugs to Linux 
	kernel developers, and 32 of them have been confirmed.
\end{itemize}

### 2023-08-02 2.motivation.tex
Finish a temporary version (without comparing experiment and some evaluation results)

Figure~\ref{fig_demo_bug} shows a real null-pointer dereference caused by a 
data race in the Linux DRM driver. In the DRM driver, the functions {\tt 
exynos\_drm\_crtc\_atomic\_disable()} and {\tt exynos\_crtc\_handle\_event()} 
can execute concurrently. We exploit \textcircled{\footnotesize{n}} to 
represent the execution order of instructions and show one execution case on 
the left of Figure~\ref{fig_demo_bug}. In Thread 1, the variables {\tt 
crtc->state->event} and {\tt crtc->state->active} are checked by an if 
statement in the function {\tt exynos\_drm\_crtc\_atomic\_disable()}. After the 
condition is calculated to be true, the function {\tt 
exynos\_crtc\_handle\_event()} is executed in Thread 2. In this function, the 
value of {\tt crtc->state->active} is assigned to {\tt event} 
(\textcircled{\footnotesize{2}}) and then {\tt event} is checked in an if 
statement (\textcircled{\footnotesize{3}}). If it is not NULL, the variable 
{\tt crtc->state->event} is assigned with NULL 
(\textcircled{\footnotesize{4}}). Right after this assignment, the function 
{\tt drm\_crtc\_send\_vblank\_event} is called  in Thread 1 
(\textcircled{\footnotesize{6}}) with the argument {\tt crtc->state->active}, 
after acquiring the lock {\tt crtc->dev->event\_lock} 
(\textcircled{\footnotesize{5}}). In the called function, the variable {\tt 
crtc->state->event} is dereferenced through {\tt e->pipe} 
(\textcircled{\footnotesize{6}}). In this execution case, the data structure 
{\tt crtc->state->event} is first assigned with NULL and then dereferenced, and 
thus a null-pointer dereference can occur.

@@

\PP{C1: Getting locking rules.} The relationship between variables and locks 
is not well documented in OS kernels, making it hard to determine whether a 
specific variable should be protected by a lock and which lock is 
required (locking rules), even for an expert developer. And thus existing 
annotation-based approaches~\cite{Boyapati:OOPSLA02, Anderson:PLDI08, 
Anderson:PLDI09, Zhou:MICRO19, Flanagan:PASTE01, Flanagan:PLDI00, 
Sadowski:PLATEAU14, ClangThreadSafety, Blackshear:OOPSLA18} are difficult to 
apply to race detection in OS kernels. Other approaches~\cite{Choi:PLDI02, 
Engler:SOSP03, Voung:FSE07, Pratikakis:PLDI06, Naik:PLDI06} employ 
lockset-based analysis to detect data races automatically, but they do not 
consider alias relations~\cite{Voung:FSE07, Engler:SOSP03} or just use 
imprecise flow-insensitive alias analysis~\cite{Choi:PLDI02, 	
Pratikakis:PLDI06, Naik:PLDI06}. However, due to the heavy use of pointers and 
data structure fields in OS code, the alias relationships between variables can 
be very complex, and thus lacking effective alias analysis can introduce many 
false locking rules.

\PP{C2: Dropping false data races.} Static analysis suffers from false 
positives. For example, each data race involves more than one code path that 
should be able to concurrently execute. However, which code can execute 
concurrently is not well documented for an OS kernel, and it is also hard to 
determine concurrent code statically due to the complexity of OS code. Thus 
static analysis can report many false data races.

\PP{C3: Estimating harmfulness of data races.} Many data races are benign, and 
can not cause memory or logic bugs, and thus developers are unwilling to put 
effort into repairing them. To automatically detect harmful data races, most 
approaches~\cite{Narayanasamy:PLDI07, Sen:PLDI08, Kasikci:SOSP13, 
Kasikci:ASPLOS12} estimate the harmfulness of data races through dynamic 
analysis, and explore thread interleavings to trigger data races to estimate 
their harmfulness. However, they suffer from low code coverage and thus can 
miss many real harmful data races.

>>>>

Figure~\ref{fig_demo_bug} shows a real null-pointer dereference caused by a 
data race in the Linux DRM driver. In this example, We exploit 
\textcircled{\footnotesize{n}} to represent the execution order of instructions 
and show one execution case on the left of Figure~\ref{fig_demo_bug}. In the 
DRM driver, the functions {\tt exynos\_drm\_crtc\_atomic\_disable()} and {\tt 
exynos\_crtc\_handle\_event()} can execute concurrently. In Thread 1, the 
variables {\tt crtc->state->event} and {\tt crtc->state->active} are checked by 
an if statement in the function {\tt exynos\_drm\_crtc\_atomic\_disable()}. 
After the condition is calculated to be true, the function {\tt 
exynos\_crtc\_handle\_event()} execute in Thread 2. In this function, the value 
of {\tt crtc->state->event} is assigned to {\tt event} 
(\textcircled{\footnotesize{2}}) and then {\tt event} is checked in an if 
statement (\textcircled{\footnotesize{3}}). If it is not NULL, the variable 
{\tt crtc->state->event} is assigned with NULL 
(\textcircled{\footnotesize{4}}). Right after this assignment, the function 
{\tt drm\_crtc\_send\_vblank\_event()} is called  in Thread 1 
(\textcircled{\footnotesize{6}}) with the argument {\tt crtc->state->event}, 
after acquiring the lock {\tt crtc->dev->event\_lock} 
(\textcircled{\footnotesize{5}}). In the called function, the variable {\tt 
crtc->state->event} is finally dereferenced through {\tt e->pipe} 
(\textcircled{\footnotesize{7}}). In this execution case, the variable {\tt 
crtc->state->event} is first assigned with a NULL value and then dereferenced, 
and thus a null-pointer dereference can occur.

@@

\PP{C1: Getting locking rules.} The relationship between variables and locks is 
not well documented in OS kernels, making it hard to determine whether a 
specific variable should be protected by a lock and which lock is required, 
even for an expert developer. And thus existing 
annotation-basedapproaches~\cite{Boyapati:OOPSLA02, Anderson:PLDI08, 
Anderson:PLDI09, Zhou:MICRO19, Flanagan:PASTE01, Flanagan:PLDI00, 
Sadowski:PLATEAU14, ClangThreadSafety, Blackshear:OOPSLA18} are difficult to 
apply to race detection in OS kernels. Other approaches~\cite{Choi:PLDI02, 
Engler:SOSP03, Voung:FSE07, Pratikakis:PLDI06, Naik:PLDI06} employ 
lockset-based analysis to detect data races automatically, but they do not 
consider alias relationships~\cite{Voung:FSE07, Engler:SOSP03} or just use 
imprecise flow-insensitive alias analysis~\cite{Choi:PLDI02, 	
Pratikakis:PLDI06, Naik:PLDI06}. However, due to the heavy use of pointers and 
data structure fields in kernel code, the alias relationships between variables 
can be very complex, and thus lacking effective alias analysis can introduce 
both false positives and false negatives.

\PP{C2: Dropping false data races.} Due to lack of run-time information, static 
analysis suffers from false positives. For example, each data race involves 
more than one code path that should be able to concurrently execute. However, 
which code can execute concurrently is not well documented for an OS kernel, 
and it is also hard to determine concurrent code statically due to the 
complexity of OS code. Thus static analysis can report many false data races.

\PP{C3: Estimating harmfulness of data races.} Many data races are benign or 
introduced by developers deliberately to improve the performance of OS kernels, 
and can not cause memory or logic bugs in fact. Therefore, developers are 
unwilling to put effort into repairing these data races. To automatically 
detect harmful data races, most approaches~\cite{Narayanasamy:PLDI07, 
Sen:PLDI08, Kasikci:SOSP13, Kasikci:ASPLOS12} explore thread inter-leavings 
dynamically to trigger data races and observe their impact on the checked 
software. However, they suffer from low code coverage and thus can miss many 
real harmful data races.
